---
title: "Danish Kings"
author: "Terkel Sylvester"
date: "2026-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the kings

The separators in my csv `kongeraekken.csv` are semicolons!

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
library(tidyverse)
kings1 <- read.csv("data/kongeraekke.csv")

kings2 <- read_csv("data/kongeraekke.csv")

kings3 <- read.csv2("data/kongeraekke.csv")

kings4 <- read_csv2( "data/kongeraekke.csv") # this one works ! it makes the numbers numbers and the characters characters
```
Answer: 

1. Which of these functions is a `tidyverse` function?

the tidyverse functions are read_csv and read_csv2

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.

"spec_tbl_df" "tbl_df"      "tbl"         "data.frame" 

3. How many columns does the object have when created with these different functions?

read.csv and read_csv both have one column, and both read.csv2 and read_csv2 have 8 columns, but read.csv2 makes "int"-classed columns rather than "num"-classed (as with read_csv2)

4. Show the dataset so that we can see how R interprets each column

```{r eval = TRUE}
kings <- read_csv2( "data/kongeraekke.csv")

class(kings)

ncol(kings)

glimpse(kings)

```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`.

```{r}
kings <- kings %>% 
  mutate(duration = reign_end - reign_start) %>% 
  arrange(desc(duration))
```

## Calculate the average duration of reign for all rulers

```{r}
# i have three ways of selecting the duration column:
mean(kings[["duration"]], na.rm=TRUE)
mean(kings$duration, na.rm=TRUE)
# both of these functions turn the column into a vector

kings %>% 
  summarize(mean_duration = mean(duration, na.rm=TRUE))
# this instead works on a tibble
```

## How many and which kings enjoyed a longer-than-average duration of reign?

```{r}
# creating a new tibble only including kings who reigned for a longer-than-average time:
longer_than_avg <- kings %>% 
  filter(duration>19.7)

head(longer_than_avg)
nrow(longer_than_avg) # returns 28, showing that there have been 28 kings with an above-average reign duration
```


## How many days did the three longest-ruling monarchs rule? 

```{r}
top_three <- kings %>% 
  arrange(desc(duration)) %>% 
  slice_head(n=3) %>% 
  mutate(Days = duration*365) %>% 
  select(name, duration, Days)

head(top_three)
```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

```{r}
# first, we add a midyear-column (for plotting to the x-axis)
kings <- kings %>% 
  mutate(midyear = reign_start + (reign_end-reign_start)/2)

kings %>% 
  ggplot(aes(x = midyear, y = duration)) +
  geom_point()
# shows that reigns are generally longer later, but very scattered indeed 

kings %>% 
  ggplot(aes(x = midyear, y = duration)) +
  geom_smooth()
# more of a big picture-visualisation: Christian 4.s 60 year long reign, while obvious in the above geom_point()-visualisation, has here very little effect
```
